Investigating social change during cultural contact period using geometric morphometry of pottery shapes from Iron Age northeastern Taiwan

### 2020-01-26 analysis/paper/paper.Rmd
para 4

To identify craft specialization in the archaeological record, variation in the standardization of a specialized product is often used as an indicator. This is based on the assumption that specialized mass production will lead to uniformity of the product due to routinization, higher skills, and fewer producers involved in production [@Stark1995; @Costin1991; @Costin2001]. For investigating ceramic standardization, several measurements have been discussed and used, such as metric, compositional, and technological approaches [@Arnold2000; @Boness2015; @Rice1991; @Blackman1993; @Costin1991; @Tite1999; @Roux2015]. Among those variables, metric measurements are most widely applied to archaeological assemblages because of the ease of collecting these data. By comparing variations for sets of data using coefficient of variation coupled with statistical testing, the degree of standardization can be identified according to values [@Eerkens2001; @Roux2003; @Stark1995]. @Roux2018's ethnographic study in the Jodhpur region of India suggests the number of artisans can be statistically assessed by comparing coefficients of variation and differences between artisans can be detected at both intra-individual level and inter-individual level in a region. However, linear measurements might be limited for identifying differences in variations because they could be insensitive to subtle variations resulting from changes in assemblages. Shape analysis based on geometric morphometrics method are new approaches for measuring the degree of standardization since it takes the overall shape as variable for comparison [@Slice2007]. Using geometric morphometrics approaches, we study shape and standardization of locally made ceramics at Kiwulan (1350-1950 AD) [@Chen2007], a large multi-component archaeological site in northeastern Taiwan, to identify changes in ceramic production in cross-cultural interaction contexts.

Our hypothesis is that if foreign influences had impacts on the emergence of social inequality in local indigenous society due to the monopolies for trade between a small number of indigenous people, then we expect to see the changes in social organization from a more corporate to a more network organization that would be reflected by pottery production at Kiwulan [@Feinman2000]. The emergence of craft specialization, here pottery specialization, are usually related to changes in social organization towards a society with increasing status inequality. In this case, if the competitions for foreign resources and being trade partners of European or Chinese colonizer among individuals gradually lead to increasing social inequality, then we expect the local ceramic will show more homogeneous features after contact due to the emergence of craft specialization caused by control of small group of individuals. This study is important to understand the indirect influences of the European colonists on the local indigenous societies, which remains unclear in East Asia where the colonial power was not successful compared to other places in Southeast Asia. Also, it would help to understand the relationship among the use of prestige goods, the degree of ceramic specialization, and the influence of foreign colonizers in a pericolonial context [cf. @Acabado2017].

## Geometric Morphometric approaches

>>>>

To identify craft specialization in the archaeological record, variation in the standardization of a specialized product is often used as an indicator. This is based on the assumption that specialized mass production will lead to uniformity of the product due to routinization, higher skills, and fewer producers involved in production [@Stark1995; @Costin1991; @Costin2001]. For investigating ceramic standardization, several measurements have been discussed and used, such as metric, compositional, and technological approaches [@Arnold2000; @Boness2015; @Rice1991; @Blackman1993; @Costin1991; @Tite1999; @Roux2015]. Among those variables, metric measurements are most widely applied to archaeological assemblages because of the ease of collecting these data. By comparing variations for sets of data using coefficient of variation coupled with statistical testing, the degree of standardization can be identified according to values [@Eerkens2001; @Roux2003; @Stark1995]. @Roux2018's ethnographic study in the Jodhpur region of India suggests the number of artisans can be statistically assessed by comparing coefficients of variation of linear dimensions of pots, and differences between artisans can be detected at both intra-individual level and inter-individual level in a region. However, because pottery vessels are typically have curved shapes, linear measurements have limited sensitivity to many kinds of shape variations. Thus, to capture subtle shape variations that might be relevant to standardization, a method that is senstive to non-linear aspects of shape variation is necessary. Here we analyse pottery shape using geometric morphometric methods, a new approach for measuring the degree of standardization because it takes a statistical summary of the overall shape as variable for comparison [@Slice2007]. Using geometric morphometry, we study shape and standardization of locally made ceramics at Kiwulan (1350-1950 AD) [@Chen2007], a large multi-component archaeological site in northeastern Taiwan, to identify changes in ceramic production resulting from cross-cultural interaction contexts.

Our hypothesis is that if foreign influences had impacts on the emergence of social inequality in local Indigenous society due to monopolies of trade arising among a small number of Indigenous people, then we expect to see changes in social organization from a more corporate to a more network organization [@Feinman2000]. We expect this corporate-network shift to manifest as increased specialization pottery manufacture, indicated archaeologically by icreased standardisation of pottery production at Kiwulan. The emergence of craft specialization, here pottery specialization, is often related to changes in social organization towards a society with increasing status inequality. In this case, if the competition among Indigenous individual for foreign resources and trade partnership with European or Chinese colonizers gradually lead to increasing social inequality, then we expect the local ceramics will show more homogeneous shapes after contact due to the emergence of craft specialization caused by rgreater economic and social control by a small group of individuals at Kiwulan. Our results are important for understanding the indirect influences of  European colonists on local Indigenous societies, which remains unclear in East Asia where colonial power had limited success compared to other places in Southeast Asia. Our findings improve our understanding of the relationship between the use of prestige goods, the degree of ceramic specialization, and the influence of foreign colonizers in a pericolonial context [cf. @Acabado2017].

## Geometric Morphometry
